White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-09-09 papers/wdmerger_II/paper.tex
Narrow focus to a single 2D 0.64+0.64 C/O WD problem

We implement the white dwarf collision problem in \castro\ in the \wdmerger\ problem
directory. We prepare the domain in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other along the $x$-axis.
The composition of the WDs is as described in Paper I: low-mass WDs are helium-dominated,
intermediate-mass WDs are carbon and oxygen dominated, and high-mass WDs contain
oxygen, neon, and magnesium. By default the WDs approach each other head-on, with
their respective centers lying on the $x$-axis, but we have also added an option
to have the WDs approach each other at a non-zero impact parameter $b$, where the
offset is perpendicular to the $x$-axis in the $y$ direction. We measure
$b$ in units of the secondary WD's radius, so that the default head-on case
corresponds to $b = 0$ and the case where the two WDs just graze past each other
corresponds to $b = 1$ (in practice it happens slightly differently because
of the WDs responding to each other's gravitational fields).

For all the 3D simulations to follow, we use the same coarse grid as in Paper I:
there are 256 zones per dimension, and the width of each axis is $1.024 \times 10^{10}$
cm, yielding a base resolution of 400 km. When we use adaptive-mesh refinement, we will
use the refinement scheme described in \autoref{sec:unstable_burning}. This is
acceptable because the outcome of the collision problem is chiefly a function
of the propagation of burning front through the WDs, so for this problem one
can take some liberties in other parts of the algorithm for the purpose of
computational efficiency, which can always be turned back on later if a more
accurate answer is desired. So, for example, we do not need to add refinement
everywhere in the WDs in the interest of getting the collision dynamics to be
slightly more accurate. Additionally, we choose to solve the Poisson equation
for the self-gravity of the system only on the coarse grid. The gravitational
potential and acceleration on the finer levels are interpolated from the coarse
grid. This grants enhanced computational speed without a serious effect on the
accuracy of the simulation (for the case where the fine grids cover the stars,
we have checked that the more accurate gravitational forcing would only modify
slightly the time to impact and the subsequent detonation process).

We have also added to \wdmerger\ the ability to use the 2D cylindrical ($R-z$)
coordinate system evolution in \castro. In this coordinate system, we align
the WDs along the $z$-axis (which is analogous to our $x$-axis in the Cartesian
evolution) at the usual distance, with the center of the WDs at $R = 0$. The
un-simulated $\phi$ dimension then would extend the WDs through a $2\pi$ revolution.
The nature of the axisymmetry inherent to the cylindrical coordinate system
means that we can only run head-on collisions with $b = 0$. The coarse grid
is the same resolution as the 3D case: the width of the $z$ axis is $1.024 \times 10^{10}$
cm and the width of the $R$ axis is $5.012 \times 10^{9}$ cm, with twice as
many zones along the $z$ axis as the $R$ axis so that the equal 400 km
resolution is maintained.

@@

As in Paper I, the region outside the WDs is filled with a low-density ambient
material with $\rho = 10^{-4}\ \text{g / cm}^{-3}$ whose composition is the
same as that of the WDs (which is equal carbon and oxygen by mass, unless
stated otherwise). Reactions in this region are unimportant, so for
computational efficiency, we disable all nuclear burning for zones that have
$\rho < 10^6\ \text{g / cm}^{-3}$ and $T < 10^8\ \text{K}$. The initial ambient
temperature is $10^7\ \text{K}$, which is the same as the initial temperature
everywhere throughout the WDs, and we set the temperature floor to the same
value (any zone that reaches below this temperature will be reset).

@@

In contrast, the suppressed burning mode (modeled after \citet{kushnir:2013}) produces only about
$5 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed. In order to ensure that the
very different results we obtained were not a result of the specific choice we made for how to suppress
the burn, we tried varying the approach discussed in \autoref{sec:burner}. First, we limited based on
the current value of the internal energy and sound speed rather than the fixed, initial value during a burn
(this should be a less severe limiter). This did not yield a significant difference in the outcome. Second,
we also tried applying the limiting after the burn had completed, using the differences between the input
and output states of the integration, rather than directly limiting the RHS in the ODE. The behavior of the
burn was qualitatively different, in that nickel production began much earlier in the collision when the limiting
was applied this way. Nevertheless, the qualitative final outcome of the collision -- only trace amounts
of nickel produced in total -- was still unchanged. With this variant too, it did not matter whether the limiting
factor was determined based on the initial value of the state or the final value of the state.
Our default 400 km resolution is significantly lower than that of \citeauthor{kushnir:2013} (the text
says they used resolutions up to a few km) but we have tried runs at 100 km and 50 km resolution and
the results are qualitatively similar. We did not directly compare the results of the method at
the same resolution as theirs, but our results suggest at least that this limiter significantly
modifies the evolution for modest resolution simulations and may be inappropriate as a method for enabling
simulations at those resolutions to obtain accurate results.

\subsection{Impact Parameter}
\label{sec:parameters:impactparameter}

As the impact parameter varies for a head-on collision, the cross-section of the contact
zone between the two stars varies in size and geometry. In \autoref{fig:impact_parameter}, we
see the dependence of the nickel production on the impact parameter $b$, measured in units
of the WD radius. This test was performed using the 3D setup at our usual base resolution of
400 km. For computational efficiency, timestep limiting based on burning was disabled.
We sampled at $b = 0.0$, $0.01$, $0.02$, $0.03$, $0.04$, $0.05$, and then increasing
in increments of $0.05$. Relative to $b = 0$, there is a slight increase in nickel production for small
but non-zero $b$, which then declines with increasing $b$ except for a bump at $b = 0.5$.
For $b \ge 0.7$, no detonation forms at all, and the merged object remains a stable C/O WD
with $M = 1.2\ \msolar$ at least until $t = 100$.

It is worth nothing that for $b = 0.0$ the nickel production is a bit smaller for the 3D simulation
than its equivalent 2D simulation, the default case described above. There are differences both in
the size of the timestep (which as we saw in \autoref{sec:parameters:timestepping} plays a role in
the total nickel production) and in the discretized nature of the stars; the assumption of perfect
azimuthal symmetry inherent to the 2D simulation breaks down for the 3D simulation, especially at
low resolution, since the zones are rectangular.

\begin{figure}
  \centering
  \includegraphics[scale=0.425]{plots/impact_parameter}
  \caption[Nickel production dependence on the impact parameter]
          {$^{56}$Ni generation as a function of the initial WD impact parameter for
           a $0.64\ \msolar$ + $0.64\ \msolar$ WD collision.
           \label{fig:impact_parameter}}
\end{figure}

@@

We tested a few other parameters that ended up having no serious effect on the outcome.
The dual energy hydrodynamic formalism (discussed in Paper I) has parameters that determine
whether we obtain the internal energy from the total energy or from a separately advected
internal energy variable. This is usually employed when calculating the pressure during
a hydrodynamics update, and we have also added to \castro\ a similar switch when calculating
the initial temperature for a burn. We find that the value of the dual energy parameter has
only a minor effect on the nickel production, yielding variations on the order of a few percent
with no systematic trend. We also tested whether our choice for the temperature floor mattered
by lowering it to $10^6\ \text{K}$ and $10^5\ \text{K}$; this had a minor effect on nickel
production, again on the order of a few percent. It does allow parts of the WDs to
become colder as they move, due to numerical errors in the advecting flow (the Helmholtz
equation of state is very insensitive to temperature, so even small changes in the internal
energy can create very large changes in the implied temperature), but the amount of energy
injected by the collision is so large that the final result is largely insensitive to whatever
the initial temperature of the
WDs were. We also checked whether there was any consequence of our choice to disable the
burning for small enough $\rho$ and $T$. Lowering the minimum $\rho$ for reactions to
$10^0\ \text{g cm}^{-3}$ from $10^6\ \text{g cm}^{-3}$ had a negligible effect,
and varying the minimum $T$ for reactions from $2 \times 10^{7}\ \text{K}$ to
$4 \times 10^{8}\ \text{K}$ also had no meaningful effect.

@@

The choice of the WD composition to be equal carbon and oxygen by mass is a common
choice in the literature and is broadly comparable to what stellar evolution codes
yield for WDs of this mass. Still, it is an arbitrary approximation, especially the
insistence that the composition is uniform throughout the WD. In \autoref{table:co}
we list the nickel production as a function of the initial carbon mass fraction.
This can have a few percent impact on the outcome for the plausible values we tested.
In future work we may look at the effect of a more realistic progenitor structure,
including the effects of a helium shell on the surface of the WDs \citep{holcomb:2015}
and a non-uniform interior structure.

\input{plots/co_frac_m_P_0.64_m_S_0.64.tbl}

>>>>

We implement the white dwarf collision problem in \castro\ in the \wdmerger\ problem
directory. We prepare the domain in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other along a coordinate axis.
In this paper we focus only on head-on collisions of two 0.64 $\msolar$ white dwarfs
composed of equal parts carbon and oxygen by mass, for comparison to earlier studies;
however, the code does have the ability to vary the impact parameter of the collision
and the composition of the WDs.

We have added to the \wdmerger\ problem setup the ability to take advantage of 
the 2D cylindrical ($R-z$) coordinate system evolution in \castro, and we use it
in this study, because the axisymmetry inherent to the cylindrical coordinate system
lends itself well to the head-on collision problem. In this coordinate system, we
align the WDs along the $z$-axis (which is analogous to the $x$-axis in Cartesian
evolution), with the center of the WDs at $R = 0$. The un-simulated $\phi$ dimension
then would extend the WDs through a $2\pi$ revolution. The domain has width $4 \times 10^{9}$
cm along the $z$ axis, and is half as wide along the $R$ axis. The resolution is equal
in both dimensions, so that there are twice as many zones in the $z$ dimension in
a uniform grid. The default grid has 256 zones along the $z$ axis, yielding a resolution
of 156.25 km.

@@

As in Paper I, the region outside the WDs is filled with a low-density ambient
material with $\rho = 10^{-4}\ \text{g / cm}^{-3}$ whose composition is the
same as that of the WDs. Reactions in this region are unimportant, so for
computational efficiency, we disable all nuclear burning for zones that have
$\rho < 10^6\ \text{g / cm}^{-3}$ and $T < 10^8\ \text{K}$. The initial ambient
temperature is $10^7\ \text{K}$, which is the same as the initial temperature
everywhere throughout the WDs, and we set the temperature floor to the same
value (any zone that reaches below this temperature will be reset).

@@

In contrast, the suppressed burning mode (modeled after \citet{kushnir:2013}) produces only about
$5 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed. In order to ensure that the
very different results we obtained were not a result of the specific choice we made for how to suppress
the burn, we tried varying the approach discussed in \autoref{sec:burner}. First, we limited based on
the current value of the internal energy and sound speed rather than the fixed, initial value during a burn
(this should be a less severe limiter). This did not yield a significant difference in the outcome. Second,
we also tried applying the limiting after the burn had completed, using the differences between the input
and output states of the integration, rather than directly limiting the RHS in the ODE. The behavior of the
burn was qualitatively different, in that nickel production began much earlier in the collision when the limiting
was applied this way. Nevertheless, the qualitative final outcome of the collision -- only trace amounts
of nickel produced in total -- was still unchanged. With this variant too, it did not matter whether the limiting
factor was determined based on the initial value of the state or the final value of the state.
Our default resolution is significantly lower than that of \citeauthor{kushnir:2013} (the text
says they used resolutions up to a few km) but we have tried runs at higher resolution and
the results are qualitatively similar. We did not directly compare the results of the method at
the same resolution as theirs, but our results suggest at least that this limiter significantly
modifies the evolution for modest resolution simulations and may be inappropriate as a method for enabling
simulations at those resolutions to obtain accurate results.

@@

We tested a few other parameters that ended up having no serious effect on the outcome.
The dual energy hydrodynamic formalism (discussed in Paper I) has parameters that determine
whether we obtain the internal energy from the total energy or from a separately advected
internal energy variable. This is usually employed when calculating the pressure during
a hydrodynamics update, and we have also added to \castro\ a similar switch when calculating
the initial temperature for a burn. We find that the value of the dual energy parameter has
only a minor effect on the nickel production, yielding variations on the order of a few percent
with no systematic trend. We also tested whether our choice for the temperature floor mattered
by lowering it to $10^6\ \text{K}$ and $10^5\ \text{K}$; this had a minor effect on nickel
production, again on the order of a few percent. It does allow parts of the WDs to
become colder as they move, due to numerical errors in the advecting flow (the Helmholtz
equation of state is very insensitive to temperature, so even small changes in the internal
energy can create very large changes in the implied temperature), but the amount of energy
injected by the collision is so large that the final result is largely insensitive to whatever
the initial temperature of the
WDs were. We also checked whether there was any consequence of our choice to disable the
burning for small enough $\rho$ and $T$. Lowering the minimum $\rho$ for reactions to
$10^0\ \text{g cm}^{-3}$ from $10^6\ \text{g cm}^{-3}$ had a negligible effect,
and varying the minimum $T$ for reactions from $2 \times 10^{7}\ \text{K}$ to
$4 \times 10^{8}\ \text{K}$ also had no meaningful effect. Finally, we also checked
that the 2D axisymmetric simulation matched the results of an equivalent 3D Cartesian simulation.

@@
